### 2020-07-10 case-studies.tex
finally figured out how to formalize closure collection properly... subtle

>>>>

We assume that the user has selected at most one closure for each livelit in the unexpanded expression.
The corresponding environment is applied as a substitution before attempting to evaluate the requested expression.
To maintain the context independence principle, this request must take the form of a closed function 
of the splices. 
As discussed in Sec.~\ref{sec:live-evaluation}, evaluation can fail if the free variables in the splices 
overlap with the free variables in selected closure.

### 2020-07-10 livelit-calculus.tex
finally figured out how to formalize closure collection properly... subtle

\subsection{Live Feedback via Closure Collection}
In order to support live feedback, a livelit needs to be able to ask the system
to evaluate expressions under a selected closure for the hole that the livelit is filling.
This mechanism was described by example in Sec.~\ref{sec:image-transformation}.

In this section, we will formalize the process of efficiently collecting closures 
for the livelits that appear in a program. The key idea is that we generate a temporary expansion,
called the \emph{cc-expansion}, 
where each livelit expands to an empty hole applied to each of its splices. 
The key rule for the cc-expansion judgement, $\ccexpands{\Phi}{\Gamma}{\hat e}{e}{\tau}$, is:
\begin{mathpar}
\inferrule[CCApLivelit]{
    \livelitCtxEntry{a}{\taut{expand}}{\taut{model}}{\dtxt{expand}} \in \Phi\\\\
    \{ \ccexpands{\Phi}{\Gamma}{\hat e_i}{e_i}{\tau_i} \}_{i < n}
}{
    \ccexpands{\Phi}{\Gamma}{\haplivelit{u}{a}{\dtxt{model}}{\splat{\hsplice{\hat e_i}{\tau_i}}}}{\hap{\hehole{u}}{\splat{e_i}}}{\taut{expand}}
}
\end{mathpar}
We then elaborate and evaluate the cc-expansion 
following Hazelnut Live's dynamic semantics for programs with holes, which were summarized 
in Sec.~\ref{sec:external-and-internal-lang}. The result will contain some number of hole closures
for each livelit hole. We call these the proto-closures and their associated environments the proto-environments for that livelit hole, formally defined as follows.
\begin{definition}[Proto-Environment Collection]
If $\ccexpands{\Phi}{\cdot}{\hat e}{e}{\tau}$ and $\elabs{}{e}{d}{\tau}{\Delta}$
and $\evalsTo{d}{d'}$ then $\protoEnvsOf{\hat e}{u} = \{ \sigma \mid \dehole{u}{\sigma} \in d' \}$.
    % protoclosures(\hat e, u) = { sigma | hole^u_sigma in d' } where \hat e cc-expands to e and e elaborates to d and d evaluates to d' 
\end{definition}
A proto-environment for a livelit hole might itself contain a proto-closure for another livelit hole.
For example, in Fig.~\ref{fig:color}, the proto-environment for the \li{\$color} livelit contained the 
proto-closure for the \li{\$slider} livelit because it was used in the value of a bound variable, \li{grayLevel}.
This is unsatisfying, because it means that the actual value of that variable, which requires computing the expansion,
 is not available to the \li{\$color}
livelit as expected.
Consequently, we must fill any livelit holes that appear in the proto-environments for other livelit holes.
We do so by computing the parameterized expansions of the corresponding livelits,  
as specified in Sec.~\ref{sec:calculus-expansion} (we can skip the final premise of Rule \rulename{EApLivelit} because the 
splice arguments have already been applied in the cc-expansion). We then perform a hole filling operation, 
$\instantiate{\dtxt{pexpansion}}{u}{d}$ over each of the proto-environments associated with livelit holes. 
Hole filling was defined by Hazelnut Live and operates to 
fill every closure for hole $u$ in $d'$ with $\dtxt{pexpansion}$ and derives from the metavariable
instantiation operation of contextual modal type theory (CMTT) \cite{HazelnutLive,Nanevski2008}. 
Unlike substitution, hole filling is
not capture-avoiding. Instead, the environment on each of these closures is applied to $\dtxt{pexpansion}$
as a substitution, i.e. the delayed substitutions captured in the environment are realized.
In this case, however, the proto-expansion is guaranteed to be closed due to 
the context independence discipline we maintain in Rule \rulename{EApLivelit}, 
so hole filling acts simply as a syntactic replacement.

Let $\livelitsIn{\hat e}$ by the set of hole names for the livelits that appear in $\hat{e}$.

\begin{definition}[Environment Collection]
    If $u \in \livelitsIn{\hat e}$ and $\protoEnvsOf{\hat e}{u} = \splat{\sigma_i}$ 
    % then $\EnvsOf{\hat e}{u} = \splat{fillem{\hat e\sigma_i}}$.
\end{definition}

This is not sufficient, however, because some of these hole closures might themselves 
contain livelit holes. However, a livelit that uses this closure needs the final result, 
not a hole. Consequently, we need to \emph{fill and resume} livelit closures that appear inside 
other livelit closures. Fortunately, for a pure functional language can do so efficiently 
by computing the expansion for the corresponding livelit expression, elaborating it in isolation,
filling the closure with the resulting elaboration, and then resuming evaluation. A theorem from
the prior paper establishes the correctness of this process, i.e. that the resulting program will 
produce the same result as if it had been evaluated from scratch.

In fact, even the cc-expansion evaluation process does not go to waste -- we can resume the 
result of cc-expansion evaluation in the same way to arrive at the same final result 
as the standard expansion and evaluation process.

>>>>

\subsection{Live Feedback via Closure Collection}
In order to support live feedback, a livelit needs to be able to ask the system
to evaluate expressions under one of the closures associated with the livelit.
This mechanism was introduced by example in Sec.~\ref{sec:image-transformation}.
In this section, we will formalize the process of efficiently collecting closures 
for the livelits that appear in a program.

The key idea is that we generate an alternative expansion,
called the \emph{cc-expansion}, 
where each livelit expands to an empty hole applied to the splices. In other words,
we leave a hole in place of the parameterized expansion, which we expand and record in the \emph{cc-context}, $\Omega$, on the side.
The key rule for the cc-expansion judgement, $\ccexpands{\Phi}{\Gamma}{\hat e}{e}{\tau}{\Omega}$, is:
\begin{mathpar}
\inferrule[CCApLivelit]{
    \{ \ccexpands{\Phi}{\Gamma}{\hat e_i}{e_i}{\tau_i}{\Omega_i} \}_{i < n}\\
    \expands{\Phi}{\Gamma}{\haplivelit{u}{a}{\dtxt{model}}{\splat{\hsplice{\hat e_i}{\tau_i}}}}{
        \hap{\etxt{pexpansion}}{\splat{e_i}}
    }{\taut{expand}}\\
    \elabs{ }{\etxt{pexpansion}}{\dtxt{pexpansion}}{\tarr{\splat{\tau_i}}{\taut{expand}}}{\Delta}
}{
    \ccexpands{\Phi}{\Gamma}{\haplivelit{u}{a}{\dtxt{model}}{\splat{\hsplice{\hat e_i}{\tau_i}}}}{\hap{\hehole{u}}{\splat{e_i}}}{\taut{expand}}
    {\cup_{i<n} \Omega_i \cup \{\Omegaitem{u}{\dtxt{pexpansion}}\}}
}
\end{mathpar}

We then elaborate and evaluate the cc-expansion 
following Hazelnut Live's dynamic semantics for programs with holes, which were summarized 
in Sec.~\ref{sec:external-and-internal-lang}. The result will contain some number of hole closures
for each livelit hole. We call these the proto-closures and their associated environments the proto-environments for that livelit hole, formally defined as follows.
\begin{definition}[Proto-Environment Collection]
If $\ccexpands{\Phi}{\cdot}{\hat e}{e}{\tau}{\Omega}$ and $\elabs{}{e}{d}{\tau}{\Delta}$
and $\evalsTo{d}{d'}$ and $u \in \domof{\Omega}$ then $\protoEnvsOf{\Phi}{\hat e}{u} = \{ \sigma \mid \dehole{u}{\sigma} \in d' \}$.
    % protoclosures(\hat e, u) = { sigma | hole^u_sigma in d' } where \hat e cc-expands to e and e elaborates to d and d evaluates to d' 
\end{definition}
A proto-environment for a livelit hole might itself contain a proto-closure for another livelit hole.
For example, in Fig.~\ref{fig:color}, the proto-environment for the \li{\$color} livelit contained the 
proto-closure for the \li{\$slider} livelit, via the \li{grayLevel} variable.
If we use proto-environments for live evaluation, then the actual value of that variable
would not be available (here, the color could not be displayed).

Consequently, the second step of closure selection is to fill any livelit holes that appear in the proto-environments for other livelit holes.
We do so by filling them using the parameterized expansions gathered in $\Omega$ and then resuming
evaluation where appropriate.
Formally, this involves the hole filling operation $\instantiate{d_1}{u}{d_2}$ for Hazelnut Live
(which derives from the metavariable instantation operation of contextual modal type theory \cite{HazelnutLive,Nanevski2008}).
This operation 
fills every closure for hole $u$ in $d_2$ with $d_1$. 
Unlike substitution, hole filling is
not capture-avoiding. Instead, the environment on each of these closures is applied to $\dtxt{pexpansion}$
as a substitution, i.e. the delayed substitutions captured in the environment are realized.
In this case, however, the parameterized expansion is guaranteed to be closed due to 
the context independence discipline we maintain in Rule \rulename{EApLivelit}, 
so hole filling acts simply as a syntactic replacement.

Formally, we begin by defining an operation $\fillof{\Omega}{\sigma}$ which acts on proto-environments 
to fill any livelit holes that appear.
\begin{definition}[Livelit Hole Filling] ~
    \begin{enumerate}
        \item $\fillof{\Omega}{[d_1/x_1, \ldots, d_n/x_n]} = [\fillof{\Omega}{d_1}/x_1, \ldots, \fillof{\Omega}{d_n}/x_n]$
        \item $\fillof{\Omega}{d} = \instantiateB{\dtxt{pexpansion}}{u}_{\Omegaitem{u}{\dtxt{pexpansion}} \in \Omega}{d}$
    \end{enumerate}
\end{definition}

This first step may cause certain expressions to become non-final, because the filled hole is no longer
blocking evaluation. We therefore define an operation $\resumeof{\sigma}$ that resumes evalution for all closed expressions in $\sigma$.
(Open expressions can remain because some closures appear under binders in the final result, so a substitution has not yet been recorded
for the bound variables.)
\begin{definition}[Environment Resumption] ~
    \begin{enumerate}
        \item $\resumeof{[d_1/x_1, \ldots, d_n/x_n]} = [\resumeof{d_1}/x_1, \ldots, \resumeof{d_n}/x_n]$
        \item $\resumeof{d} = d'$ if $\fvof{d} = \emptyset$ and $\evalsTo{d}{d'}$
        \item $\resumeof{d} = d$ if $\fvof{d} \neq \emptyset$
    \end{enumerate}
\end{definition}

Finally, we can produce the final set of environments by filling and resuming the proto-environments.
\begin{definition}[Environment Collection]
    If $\ccexpands{\Phi}{\cdot}{\hat e}{e}{\tau}{\Omega}$
    % and $\elabs{}{e}{d}{\tau}{\Delta}$
    % and $\evalsTo{d}{d'}$ and $u \in \domof{\Omega}$
    then \[\envsOf{\Phi}{\hat e}{u} = \{\resumeof{\fillof{\Omega}{\sigma}} \mid \sigma \in \protoEnvsOf{\Phi}{\hat e}{u}\}\]
\end{definition}

This same fill and resume operation can be used to avoid recomputation when evaluating the fully expanded version of the user's program.
If the editor has already performed environment collection, then it can simply continue from where it left off
by filling and resuming
the remaining top-level livelit holes (those that do not appear in a proto-environment).

The correctness of the mechanisms described in this section rest on the fact that evaluation commutes with hole filling. 
This only holds when the language is pure: evaluation order does not matter 
so the system is free to evaluate around holes initially, 
and then return to finish the job once they have been filled.

\begin{theorem}[Post-Collection Resumption]
    If $\ccexpands{\Phi}{\cdot}{\hat e}{\etxt{cc}}{\tau}{\Omega}$ and $\elabs{}{\etxt{cc}}{\dtxt{cc}}{\tau}{\Delta}$
    and $\evalsTo{\dtxt{cc}}{\dtxt{cc}'}$ and $\resumeof{\fillof{\Omega}{\dtxt{cc}'}} = d_1$ 
    and $\expands{\Phi}{\cdot}{\hat e}{\etxt{full}}{\tau}$
    and $\elabs{}{\etxt{full}}{\dtxt{full}}{\tau}{\Delta}$ 
    and $\evalsTo{\dtxt{full}}{d_2}$ then $d_1 = d_2$.
\end{theorem}
\begin{proof}
    The key observation is that filling the livelit holes in the cc-expansion gives the full expansion,
    i.e. $\fillof{\Omega}{\dtxt{cc}} = \dtxt{full}$. Resumption is simply evaluation for closed expressions.
    By commutativity of hole filling, established in the prior work \cite{HazelnutLive}, 
    we can delay hole filling until $\dtxt{cc}$ has first been evaluated to
    $\dtxt{cc}'$.
\end{proof}

In a language with side effects, one would instead need to modify evaluation of  
cc-expansions such that the expansion of each livelit application is evaluated alongside the 
temporary livelit expansions,
with the result stored for subsequent hole filling. This would ensure that
side effects happen in the same order and only once. The details are beyond the scope of this work.
